Literature Review

### 2019-05-12 background.Rmd
updates

Undergraduate introductory STEM courses often include technical writing training. Too often it is treated as an incidental skill that each student develops "along the way." At other times writing instruction is haphazardly designed, implemented, and executed. Adding to the learning challenge, faculty may not be the ones helping students develop writing skills. If technical writing training is part of a lab course, graduate teaching assistants or even advanced undergraduates may be who teaches technical writing. TAs may have under-developed writing skills themselves, lack pedagogical content knowledge needed to guide students effectively, or not have been explicitly trained to teach writing in a particular way.

The essential first step towards improving the current situation is widespread adoption of evidence-based writing training. Equally important is reducing the practical barriers that keep faculty from routinely including scientific writing in post-secondary STEM classes.

In 2014 we launched a report data intake process we call **SAWHET** that systematically collects writing and other data from student writers and graduate instructors. Using these data we are mapping key features of student writing, as well as connections and correlations between those features. These insights then inform how we teach technical writing and train graduate instructors.

Currently our structured database contains >5000 laboratory reports written by undergraduate students for introductory biology courses, all in standardized .docx and .txt formats. Additional data and metadata available for each report includes:

@@

To that end, our program trains graduate TAs to use a **global issues / biggest issues first** approach to correcting undergraduate technical writing. Ideally:

*  TAs spend a shorter amount of time writing fewer comments that focus on broader issues.
*  TAs provide no more than 3-5 comments per text page.
*  Comments encourage student reflection and self-correction rather than prescribe specific changes. 
*  As students' writing improves, TAs shift their focus to the next most-significant errors or issues.

@@

This holistic, coaching-oriented approach is well-supported by prior research and is a recommended best practice by (among others) the Writing Across the Curriculum Clearinghouse. Yet we are uncertain how accurately TAs implement the recommended strategy. We have repeatedly asked ourselves, **"what do TAs ACTUALLY spend most of their time and effort making comments about when they grade student writing?"** Self-reported TA effort is likely to be unreliable. What is needed is a more direct measure of TA effort.

@@

* __Subject__: did the comment focus mainly on **Basic Criteria** (minimum pass/fail requirements), **Technical Issues, Writing Quality, Logic and Thinking,** or **Other** issues
* __Structure__: did the comment **Point** to an error without providing other help, provide simple **Copy Correction**, provide declarative **General or Specific Information** only, or provide **Holistic Coaching** that fosters student thinking and long-term improvement?

>>>>

Undergraduate introductory STEM courses often include technical writing training. Often it is secondary to content coverage, and so treated as an incidental skill that each student develops "along the way." Alternatively, the writing instruction process may be haphazardly designed, implemented, and executed. Adding to the learning challenge, faculty may not be who is helping students develop writing skills. Often graduate teaching assistants or even advanced undergraduates are who teaches technical writing. TAs may have under-developed writing skills themselves, lack pedagogical content knowledge needed to guide students effectively, or not have been explicitly trained to teach writing in a particular way.

The first step towards improving writing training is widespread adoption of evidence-based strategies. Equally important is reducing the practical barriers that keep faculty from routinely including scientific writing in post-secondary STEM classes.

In 2014 we launched the **SAWHET** lab report collection system, to systematically collect report documents and metadata from student writers and graduate instructors. Using these data we are mapping key features of student writing, and connections and correlations between those features. These insights are informing both how we teach technical writing and how we train graduate instructors.

Currently our structured database contains >5000 laboratory reports written by undergraduate students for introductory biology courses in standardized .docx and .txt formats. Additional data and metadata available for each report includes:

@@

To that end, we train graduate TAs to comment on undergraduate technical writing by addressing **global issues / biggest issues first**. Ideally, TAs will provide feedback within these parameters:

*  Spend less time writing fewer but more substantive comments, and provide at least two separate rounds of feedback on each document. 
    +  In routine practice, limit feedback to 3-5 comments per text page. Students cannot process more than this easily.
    +  Make recommendations that would result in the greatest improvement first.
    +  As students' writing improves, TAs can shift their focus to the next most-significant errors or issues.
*  Provide comments that encourage student reflection and self-correction rather than prescribe specific changes. 
    +  Students tend to focus on the smallest, easiest corrections first, and avoid larger, more complex revisions.
    +  Eliminating copy corrections leaves students with only the complex revisions to work on.

@@

This holistic, coaching-oriented approach is well-supported by prior research and is a recommended best practice by (among others) the Writing Across the Curriculum Clearinghouse. Yet we are uncertain how accurately TAs implement the recommended strategy. We have repeatedly asked ourselves:

**"What do TAs ACTUALLY spend most of their time and effort making comments about when they grade student writing?"**

Self-reported TA effort is likely to be unreliable. What is needed is a more direct measure of TA effort.

@@

* __Subject__: did the comment focus mainly on **Basic Criteria** (minimum pass/fail requirements), **Technical Issues, Writing Quality, Logic and Thinking,** or **Other** issues
* __Structure__: did the comment simply **Point** to an error without providing other help, provide only **Copy Correction**, provide declarative **General or Specific Information**, or provide **Holistic Coaching** that fosters student thinking and long-term improvement?

### 2019-05-12 codebook.rmd
update
updates

---
title: "The TA Comments Codebook"
date: "2/23/2019"
bibliography: library.bib
biblio-style: apalike
---

@@

#Subject Category
Is primary emphasis of the comment on:

*  Fundamental pieces
*  How to communicate generally?
*  How to collect, analyze and present data correctly?
*  The underlying rationale for the report/experiment?
*  Other issues?

Below are the choices for identifying the subject of the comment. Criteria for basic, technical, and writing choices are the same as used in the report grading rubric. To simplify sorting, administrative tags used to mark comments for further processing are included in this category.

@@

#Structure Category
Is the comment idiomatic or informative? Is there general or specific information contained in comment? Is it instructional only, or does the comment foster broader thinking?

@@

*  ??
*  Misspelling, typo, format, check, etc.
*  *Repeated phrase. (Only points to immediate error; compare to below)

@@

####Contingency Table For Holistic Scores

Scores need to have BOTH information and process (what and why) to be holistic

@@

*  Always use correct format for scientific names, i.e., italicize or underline. 
(Idiomatic because not required in our grading scheme. Incidentally, this is not accurate form either, which is one answer to ongoing arguments.)
*  You provided both scientific and common names for your plant species, so can use common names throughout the rest of the report including the title and abstract. Revise accordingly. (Idiomatic because this is not standard to all fields)

@@

There are two questions in the comment, which suggests that it is a holistic comment to pull knowledge out through thinking. However, the TA comment provides some interpretation in the next sentence. So this comment is seeking or providing specific information, while also directly fostering student’s deeper thinking. It would be tagged as “Holistic."

>>>>

---
title: "The TA Comments Codebook"
date: "`r format(Sys.time(), '%b %d, %Y')`"
bibliography: library.bib
biblio-style: apalike
---

@@

#Subject Category
Does the comment focus on or emphasize:

*  Fundamental required components?
*  How to communicate generally?
*  How to collect, analyze and present data correctly?
*  The underlying rationale for the report/experiment?
*  Other issues?

Final subject categories for comments are outlined below. Criteria for basic, technical, and writing choices are the same as used in the report grading rubric. Administrative tags used to mark comments for further processing are included in this category.

@@

#Structure Category
Is the comment idiomatic or informative? Is there general or specific information contained in the comment? Is it instructional only, or does the comment foster broader thinking?

@@

*  ??
*  One-word comments: misspelling, typo, format, check, etc.
*  Simple phrases. (Only point to immediate error; compare to below)

@@

####Contingency Table For Deciding Whether to Assign Holistic Tag

A comment needs to have BOTH information and process (what and why) to be holistic.

@@

*  Always use correct format for scientific names, i.e., italicize or underline. 
(Idiomatic because this not required in our grading scheme. Incidentally, this is not accurate form either, which is one answer to ongoing arguments.)
*  You provided both scientific and common names for your plant species, so can use common names throughout the rest of the report including the title and abstract. Revise accordingly. (Idiomatic because this is not standard to all fields, **and** is not a high-priority issue in our grading scheme.)

@@

There are two questions in the comment, which suggests that it is a holistic comment to pull knowledge out through thinking. The comment provides some interpretation in the next sentence, but is not just prescriptive. This comment is seeking or providing specific information, while also directly asking student to engage in deeper thinking. It would be tagged as “Holistic."

### 2019-05-12 findings.Rmd
update
updates

---
title: "Findings"
author: "Dan Johnson"
date: "2/22/2018"
output: html_document
bibliography: library.bib
biblio-style: apalike
---
```{r, child="_setup.Rmd"}
```

##Classifier Accuracy
The tables below show observed frequencies for 9,340 TA comments coded by hand, and frequencies for the same TA comments categorized using the optimized Naive Bayes classifier **without** pre-screening. The far right columns are overall subject frequencies, the bottom rows, overall structure frequencies.

@@

##Data Story 2

@@

#Tool or Analysis 2

@@

##Data Story 3

@@

##Data Story 4

@@

>>>>

---
title: "Findings"
output: html_document
bibliography: library.bib
biblio-style: apalike
---
```{r, child="_setup.Rmd"}
```

*Updated `r format(Sys.time(), '%b %d, %Y')`*

#Vocabulary and Word Use Patterns
##Results of Single Word Frequencies
Pearson correlations for word use frequencies compares absolute use frequencies for words found in ALL THREE of the Subject sub-categories (excluding Basic Criteria). Results of one analysis are illustrated below.

![*XY plot of relative word use frequency in technical comments vs. logic comments. *](/Users/danjohnson/Dropbox/Coding_Tools/R_Environment/R_Projects/default_website/docs/initial_exploration_files/figure-html/unnamed-chunk-3-6.png)

Results are summarized below.

Comparison|Correlation|95% CI|p-value
----------|-----------|------|-------
Writing vs. Technical|0.404|0.355 - 0.451|1.64e-47
Writing vs. Logic|0.559|0.514 - 0.602|3.60e-79
Logic vs. Technical|0.394|0.343 - 0.443|3.05e-42

The comparatively low correlations of word frequencies between subject categories suggests that some terms can serve as diagnostic keywords for particular subjects.

##Results of Log-Odds Pairs for Subject Sub-Categories
Word frequency differences between sub-categories were calculated using log-odds ratios. One example analysis is shown below.

![*Log-Odds ratio comparison of words most different between Writing comments (top) and Technical comments (bottom). *](/Users/danjohnson/Dropbox/Coding_Tools/R_Environment/R_Projects/default_website/docs/initial_exploration_files/figure-html/unnamed-chunk-7-1.png)

Words identified by individual analyses were collated into two lists: highly predictive (those appearing in two comparisions) and moderately predictive (those found in one comparison).

###Words for Subject = 2. Writing 
__Highly predictive__
Broad, broader, broadly; colloquial; language; quote, quotes, quotations; revise; tense

__Moderately predictive__
Abstract; avoid; belongs; concise; concluding; details; formal; grammar; hypothesize; ideas; implications; importance; incorrect; informal; judgement; language; locate; methods; paraphrase; precise; read; recipe; redundant; repetitive; salinity; save; starting; style; takeaway; title; typos; voice; word; wrote

###Words for Subject = 3. Technical
__Highly predictive__
Average, averaged, averages; axis; bar, bars; caption, captions; deviation; figure, figures; format, formatting; graph, graphs; legend; paired; raw; represent; sd; standard; table, tables; units

__Moderately predictive__
Chloroplast; day; df; error; label; ml; properly; stat; total

###Words for Subject = 4. Logic
__Highly predictive__
Connect, connection; expect, expected; relate; shelter

__Moderately predictive__
Affecting; benefits; biological; broader; build; calculated; chloroplast; concluding; costs; density; dots; error; evolution; expand; explore; finding; fitness; gain; grow; hypothesize; ideas; implications; importance; isolated; logical; mechanisms; prediction; procedural; relative; relevance; resources; salinity; sample; selection; sodium; temp; tie; weren't

##Results of N-Gram Frequency Comparisons
I also tried a limited scale test of an alternative approach, using 2-word phrases (2-grams) instead of single words to classify TA comments. Lists of 2-grams and relative frequencies of appearance were calculated for TA comments in all Subject categories (Basic, Writing, Technical, Logic). N-gram lists were sorted by appearance frequencies, and truncated at a within-subject frequency of <1%, producing a starting dataset of ~550 2-grams.

All 2-grams occurring at similar rates in 3 out of 4 Subject categories were classified as too general and deleted, leaving a working dataset of ~338 2-grams (out of original 550). Remaining 2-grams were tagged as potential identifiers of 1-2 Subject areas. The individual 2-gram lists were merged back to the master set and sorted alphabetically.

The compiled set was visually scanned to identify individual 2-grams and 2-gram patterns that:

* Appeared in at least 1% of all comments in one sub-group
* Appeared at a 10-fold greater frequency in one sub-group versus all others.

Full analysis of the dataset requires writing a coded search algorithm; for now, I am treating these results as s "proof-of-concept" demonstration only.

This is a partial list of the potentially useful 2-grams identified so far.

***
*  Subject = Basic
    + "basic criteria/criterion" (present in 25.12% of all Basic comments)
    + "missing outside"" (6.64% of all Basic comments)
    + "outside source(s)"" (13.75% of all Basic comments)
    + "criteria __"
    + Single words (1-grams) strongly correlated with Basic category: basic, credit, grade, criteria, fail/failing, outside
*  Subject = Writing
    + "abstract __"
    + "avoid __"
    + "capitalize __" 
    + "colloquial __"
*  Subject = Technical
    + amplitude
    + analysis/analyze
    + "a figure/graph/legend"
    + "a paired"
    + alpha level/value"
    + "annotation exercise"
    + "anova __"
    + "average __"
    + "bar(s) __"
    + "caption(s) __"
*  Subject = Logic
    + "a biological"
    + "affect __"
    + "aggression/aggressive __"
    + "allocation __"
    + "axis __"
    + "betta(s) __" (Is name of model system typically logic?)
    + big(ger) picture/concept is logic
    + connect "__"
*  N-grams Identifying TWO Subjects
    + "a citation" is basic or technical
    + "a significant" is logic or technical
    + "about __" is logic, possibly writing. Weaker example.
    + "background __" is writing or logic
    + "because that's" is writing/technical
    + "(bio)core guide/resource" is technical or writing
    + "biological __" is logic or writing
    + "broader __" is logic or writing
    + "data __" is most often technical, but occurs in others.

#Classifier Accuracy
The tables below show observed frequencies for 9,340 TA comments coded by hand, and frequencies for the same TA comments categorized using the optimized Naive Bayes classifier **without** pre-screening. The far right columns are overall subject frequencies, the bottom rows, overall structure frequencies.

@@

##Test of Horizontal Stacked Naive Bayes
One strategy to refine Naive Bayes was running the same classifier with different randomized datasets, then putting individual outputs side by side and assigning final categories based on most frequent classification in the set.

To test this I ran 5 Subject classifier replicates using the code in *Naive_bayes_V2r_2b.rmd*, with 5 different ```set.seed``` values. The exported CSV files were combined to get a set of data with up to 5 classifier assignments for the same comment.

Using multiple randomized NB training runs does not seem to help. First, it was rare for the  randomized runs to assign different incorrect classes. Second, when I compared 5 independently randomized runs by eye, incorrectly categorized items were repeatedly put into the SAME incorrect group each time. Given this, an Ensemble method of H-stacked categorical NB comparison seems unlikely to reduce overall errors in classification.

Looking at the joined data DID provide other helpful insights into the overall classification procedure though. Naive Bayes does not seem to do well with comments that:

*  Contain many or mostly punctuation symbols, such as:
    +  : not ,
    +  ?
    +  ??
    +  ???
*  Are whole numbers
    +  1
    +  2
    +  10
*  Are one word questions or comments
    +  because?
    +  "and"
*	 Are sentences enclosed in punctuation
*  Are made up entirely of STOP/common words (I am not as sure on this one)

@@

##Overall Accuracy
I looked at OVERALL accuracy of NB by recreating the 2-dimensional contingency table that I made when hand-coding, and comparing it side-by-side to NB classifications. From this I gained additional insights. Results are summarized below.

@@

![alt text](/Users/danjohnson/Dropbox/Coding_Tools/R_Environment/R_Projects/default_website/docs/NB_2r2c_files/figure-html/unnamed-chunk-4-1.png)

![alt text](/Users/danjohnson/Dropbox/Coding_Tools/R_Environment/R_Projects/default_website/docs/NB_2r2c_files/figure-html/unnamed-chunk-6-1.png)

@@

OVERALL, the 2-D Naive Bayes model performed very well for classifying comment **Subject**. When I re-ran the full original dataset (9340 comments) through the prediction model, NB estimates for Subject were within 1.5% of my hand-scored assignments. Structure classification was not as robust. Other observations:

*  NB over-classified comments as "copyediting of technical issues" by 8-10%, and under-classified comments that should be marked as "specific technical issues" by almost same amount.
*  Comments that belonged to groups comprising <1-2% of all comments were more likely to be mis-classified. 
*  Longer text comments were mis-classified more often. This suggests two follow-up methods to evaluate: 
    +  Splitting longer comments into shorter phrases
    +  Evaluating only the first 5-10 words of a comment

It appears that Subject categories are assigned mainly on the terminology used, while Structure categories are assigned based more on the sentiment and sentence structure of a comment. This suggests that a sentiment classifier or PoS tagger step needs to be added.

@@

#Looking Ahead
##Mixed Ensemble Method
Based on the results outlined above, it appears that maximum classification accuracy will be a mixed vertical Ensemble method. Going forward, the working strategy will be to:

*  Pre-screen extracted TA feedback for "short-string" comments consisting of single words, two-word phrases, more punctuation than text, etc. These represent mostly pointers and copy-editing.
*  Screen comments for 2-grams with high frequencies in one particular category. (Ex., "basic criteria" 2-gram.) 
*  Use current Naive Bayes classifier to identify Subject.
*  Evaluate whether a part-of-speech tagger can identify Structure of comments with higher accuracy, either alone when joined with NB preliminary classification.

### 2019-05-12 futurefaculty.Rmd
update
updates

A major barrier to routine writing in STEM courses is instructor knowledge. Many instructors are unaware that there are publicly available, evidence-based writing training models; instead of using these validated tools, they create their own materials from scratch. similarly, many college STEM teachers have never been introduced to proven writing training strategies, nor have they been trained to give effective, context-specific feedback [@Coil2010].

@@

####\  
***
*Works Cited*

>>>>

A major barrier to routine writing in STEM courses is instructor knowledge. Many instructors are unaware that evidence-based writing training models are available, so create their own materials from scratch. Similarly, many college STEM teachers have never been introduced to proven writing training strategies, nor have they been trained to give effective, growth-oriented feedback [@Coil2010].

@@

####\  
***
*Work Cited*

### 2019-05-12 index.Rmd
updates

##Overview
I am using classroom research and data science to better understand how students develop technical writing skills in large-enrollment biology courses.

My current NSF-sponsored grant asks whether we can accelerate student development as writers using a mix of scripted active instruction, automated support of writing in progress, and holistic feedback. One specific question we have is how **changing the structure and main focus of instructors' comments** affects students' writing development over time. To answer this we extract instructor comments from student reports, categorize the comments according to their subject and structure, then correlate the types of comments instructors make with student performance.

Classifying TA comments has been informative, but doing it by hand is impractical: 10-12,000 comments need to be scored EACH semester. There also is a significant risk of "coding drift" over time, reducing accuracy.

@@

#### \   
##Work Plan
The plan for this project is to:

1.  Compile an anonymized testing dataset of hand-coded TA comments from reports. 
2.  Import TA comments and relevant metadata from student reports into R as a "tidy" dataframe.
3.  Explore comment data structure, n-gram frequencies, etc., and identify potential elements for feature engineering.
4.  Write a supervised text classification workflow that uses Naive Bayes to assign TA comments to our pre-defined categories. 
5.  Test permutations of analysis parameters to optimize the classifier and establish baseline accuracy.
6.  Apply the optimized NB classifier to the original FULL comment dataset, identify which comments are being classified incorrectly, and look for potential patterns.
7.  Based on features identified in Steps #3 and #6, write a small set of rule- or REGEX pattern-based searches that can identify comments that are more likely to be classified incorrectly.

@@

8.  Combine the rules/pattern-based pre-screening process with the optimized Naive Bayes classifier to create a vertically  Ensemble Classifier.
9.  Re-validate the Ensemble Classifier against a subset of ~2000 independently hand-coded TA comments extracted from student lab reports for a different semester than the initial test dataset.
10. Begin using the classifier to monitor TA comment patterns in the ongoing NSF-funded project.

>>>>

##Overview
The STEM Writing Project is using classroom research and data science to better understand how students develop technical writing skills in large-enrollment biology courses. Our current NSF-sponsored grant asks if student development as writers can be accelerated using a mix of scripted active instruction, automated support of writing in progress, and holistic feedback.

One specific question we have is **does the structure and main focus of instructors' comments impact the rate of students' writing development over time?** To answer this we must extract instructor comments from student reports, categorize the comments according to their subject and structure, then correlate the types of comments instructors make with student performance.

Classifying TA comments has been informative, but doing it by hand is impractical. First, 10-12,000 comments must be scored EACH semester. Second, there is a significant risk of "coding drift" over time, reducing accuracy.

@@

#### \   
##Outline of Work Plan
The plan is to:

1.  Compile an anonymized testing dataset of hand-coded TA comments from reports. 
2.  Import TA comments and relevant metadata from student reports into R as a "tidy" dataframe.
3.  Explore comment data structure, n-gram frequencies, etc., and identify potential elements for feature engineering.
4.  Write a supervised text classification workflow that uses Naive Bayes to assign TA comments to our pre-defined categories. 
5.  Test permutations of analysis parameters to optimize the classifier and establish baseline accuracy.
6.  Apply the optimized NB classifier to the original FULL comment dataset, identify which comments are being classified incorrectly, and look for potential patterns.

@@

7.  Based on features identified in Steps #3 and #6, write a small set of rule- or REGEX pattern-based searches that can identify comments that are more likely to be classified incorrectly. 
8.  Combine the rules/pattern-based pre-screening process with the optimized Naive Bayes classifier to create a vertically  Ensemble Classifier.
9.  Re-validate the Ensemble Classifier against a subset of ~2000 independently hand-coded TA comments extracted from student lab reports for a different semester than the initial test dataset.
10. Begin using the classifier to monitor TA comment patterns in the ongoing NSF-funded project.

### 2019-05-12 litreview.Rmd
update
updates

This project brings together three disparate challenges:

*  The challenges of teaching technical writing to students in STEM disciplines
*  Needs and barriers for training future faculty to teach technical writing
*  Automated text analysis and classification

>>>>

This project raises three challenges:

*  How best to teach students in STEM disciplines to write
*  How to train future faculty to teach technical writing
*  Automated text analysis and classification
